---
title: "R for Data Science 1"
author: "Matt Chana"
date: "5/28/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1.1: What you will learn
Basic cycle is Import > Tidy > (Transform <-> Visualize <-> Model) > Communicate.

* Tidy: each column is a variable, each row is an observation
* Transforming: narrowing in on observations of interest - calculating summary statistics (Tidying + Transforming = Wrangling)
* Good visualization will show unexpected things, raise new questions about the data, hint if asking wrong question
* Models are a mathematical/computational tool


* is your big data problem is a small data problem in disguise (take a subset? - how to find right subset?)

Models used for hypothesis confirmation only, not exploration.
```{r}
# install tidyverse
# install.packages("tidyverse")

# update tityverse
# tidyverse_update()

# install other relevant packages
# install.packages(c("nycflights13", "gapminder", "Lahman"))
```

Some R basics:
```{r}
# explicitly state where a function or dataset comes from 
# dplyr::mutate()
# nycflights13::flights
```

### 1.6
"reprex" stands for minimal reproducible example. Requires:

* packages loaded at top of script (good time to check make sure up to date...)
* the data. To get the data:
```{r}
# review it, find the smallest subset that reveals the problem
mtcars
```

```{r}
# take a subset
mtcars_subset <- subset(mtcars, mpg==21.0)
mtcars_subset
```

```{r}
# output the subset in a single line
dput(mtcars_subset)
```

```{r}
# now that line can be easily loaded into a dataframe
test_data_dataframe = structure(list(mpg = c(21, 21), cyl = c(6, 6), disp = c(160, 
160), hp = c(110, 110), drat = c(3.9, 3.9), wt = c(2.62, 2.875
), qsec = c(16.46, 17.02), vs = c(0, 0), am = c(1, 1), gear = c(4, 
4), carb = c(4, 4)), .Names = c("mpg", "cyl", "disp", "hp", "drat", 
"wt", "qsec", "vs", "am", "gear", "carb"), row.names = c("Mazda RX4", 
"Mazda RX4 Wag"), class = "data.frame")
test_data_dataframe
```

* Spend time ensuring code is commended, easy to read. Concise yet informative variable names, irrelevant parts to the problem are removed
* Start a fresh r session with the code and run it to test