---
title: "movies_eda"
author: "Matt Chana"
date: "6/8/2017"
output: html_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# Import and tidy:

Initial import:
```{r}
movies <- read_csv("supporting_files/movie_metadata.csv")
movies
```

Now make easier to see:
```{r}
glimpse(movies)
```
Note, There are 5,043 observations here.

Now examine:
Check for parsing errors:
```{r}
parse_errs <- problems(movies)
parse_errs
```

looks like these movies have super inflated budgets. Why? Find and examine each row:
```{r}
parse_prob <- movies[parse_errs$row,]
glimpse(parse_prob)
```

All international films - I'm betting they're showing the `budget` column all in a foreign currency. Note, Tibble framework sets all parsing errors to `NA`, so I'll disregard for now and let them be `NA`.

Finally, look at summary of data frame to understand even better:
```{r}
summary(movies)
# str(movies)
```

Things to keep an eye on from looking at summary again:

* multiple variables have `NA` instead of a value. Ones to be aware of:
  * `duration`: strange that a movie's duration is unknown. Thankfully only 15 instances.
  * `budget`: also strange if no value here, but for the 4 instances of foreign film I found, makes sense. Many more here, at 496.
  * `title_year`: each film definitely has a year it was finished, so strange that 108 instances
  * `aspect_ratio`: 329 instances. But note not really interesting variable given aspect ratio is often modified for some films (3:4 to 16:9)
  * some values may be very excessive - like in `budget`, which has a max of approximatel 2,128,000,000. I'm sure never been movie made for over two billion dollars


Is it "tidy"?:

* each variable has its own column? Of the 28 variables shown, each represents a unique type of measurement. Check.
* each observation has its own row? Yes, no values exist that branch out into their own measurements. Check.
* each value has its own cell? No, but will treat `plot_keywords` and `generes` as-is. Notes:
  * `genres` is up to five genre keywords in alphabetical order and strings restricted to a representative keyword (ex. there is SciFi but no Scifi). Can leverage this by treating each unique combination as a measurement.
  * `plot_keywords` has a small number of keywords, but nearly infinite number of combinations of any word or short phrase from human language. So leaving this as-is (but possibly using n-grams or something else)
 
Finally, are there any duplicate observations?
```{r}
dupes <- which(duplicated(movies))
dupes
movies[dupes,]
```

Yes, so create new data frame that removes all duplicate observations:
```{r}
movies <- movies[which(!duplicated(movies)),]
movies
```

Left with 4,998 rows. So in summary, checked for parsing errors, reviewed data frame for NA and extreme values, checked if tidy, removed duplicates. Good to go!

### Review/document data and its structure
So what do I have here? No documentation on Kaggle. Looked at spreadsheet and compared against web page for a given movie. My best guesses for variables and what they mean (only ones that are not obvious):

Look at all variables:
```{r}
colnames(movies)
```

* `num_critic_for_reviews`: number of IMDB movie-critic-generated star/text reviews
* `num_voted_users`: number of IMDB user-generated star-only reviews on a scale of 1-10
* `num_user_for_reviews`: number of IMDB regular-user-generated star/text reviews

Ok, data is now tidy and documented well-enough. Moving on to EDA.


# EDA (Visualize <-> Transform <-> Model)
Review data set again:
```{r}
glimpse(movies)
```

What do I think is valuable to add at this point?
I wonder what number of genres looks like:
```{r}
movies <- mutate(movies,
                  num_genres = sapply(gregexpr("[|]", movies$genres), length))
movies
```

Plot number of genres:
```{r}
ggplot(movies) +
  geom_bar(aes(x = num_genres))
```

Now plot some other things - plot all the continuous variables to see if anything sticks out:
```{r}
library(purrr)

movies %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(x = value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()
```

```{r}
library(gridExtra)
p1 <- qplot(movies$color)
p2 <- qplot(movies$num_critic_for_reviews)
p3 <- qplot(movies$director_facebook_likes)
p4 <- qplot(movies$num_voted_users)
p5 <- qplot(movies$num_user_for_reviews)
grid.arrange(p1, p2, p3, p4, p5, ncol = 3, nrow = 2)
```









```{r}
ggplot(movies2) +
  geom_bar(aes(x = genres))
```


How many unique combinations of genres exist?
```{r}
sort(unique(movies2$genres))
length(unique(movies2$genres))
```

Find max number of genres:
```{r}
max(movies2$num_genres)
```








```{r}
ggplot(data = movies) +
  geom_point(mapping = aes(x = budget, y = imdb_score, color = color))
```


Single categorical variable in bar chart:
```{r}
 ggplot(data = movies, mapping = aes(x = content_rating)) +
  geom_bar()
```


Single continuous variable in histogram:
```{r}
ggplot(data = movies, mapping = aes(x = imdb_score)) +
  geom_histogram(binwidth = .1)
```


```{r}
ggplot(data = movies, mapping = aes(x = movie_facebook_likes)) +
  geom_histogram(binwidth = 1000)
```

